Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Quasiperiodic Variables from TESS

### 2023-08-15 paper/ms.tex
[WIP] ms: add binarity section

\begin{figure*}[!tp]
	\begin{center}
		\centering
		\includegraphics[width=0.96\textwidth]{f2.pdf}
		\vspace{-0.3cm}
		\caption{
			{\bf Validation plots used to  label CQVs}.  The complete figure
			set, with one image per sector for each of the \nallcands\
      objects in Table~\ref{tab:thetable}
			is available online {\bf For internal collaboration review:
				\url{https://www.dropbox.com/scl/fo/zlj3txot4cvymfb22wewu/h?dl=0&rlkey=3ec5f9o5xewrixzfkhkdenopa}}.
			Panels are as follows.
			{\it a)}: Phase-folded light curve; gray points are raw 2-minute
			data and black points are binned to 200 points per cycle.
			{\it b)}: Phase-dispersion minimization (PDM) periodogram.
			Dotted lines show up to the 10$^{\rm th}$ harmonic and
			subharmonic.
			{\it c)}: DSS finder chart, with 1- and 2-TESS pixel radius
			circles displayed for scale.
			{\it d)}: Cleaned light curve, binned to 20-minute cadence, in
			Barycentric TESS Julian Date (BTJD).
			{\it e)}: Phase-folded light curve, binned to 100 points per
			cycle.  The gray line denotes the automated spline-fit to the
			wrapped phase-folded light curve, and small gray triangles
			denote automatically identified local minima.
			{\it f)}: Phase-folded light curve at twice the peak period.
			{\it g)}: Phase-folded light curve at half the peak period.
			{\it h)}: Phase-folded time-series within the ``background''
			aperture defined in the SPOC light curves.
			{\it i)}: Phase-folded flux-weighted centroid in the column
			direction.
			{\it j)}: Phase-folded flux-weighted centroid in the row
			direction.
			{\it k)}: Gaia DR2 color--absolute magnitude diagram.     
			The gray background denotes stars within 100\,pc.
			{\it l)}: Information from Gaia DR2, TIC8, and the automated
			dip-counting search pipeline.  ``Neighbors'', abbreviated
			``nbhr'', are listed within apparent distances of 2 TESS pixels
			if $\Delta T$$<$2.5.
			{\it m)}: BANYAN-$\Sigma$ v1.2 association probabilities, calculated
			using positions, proper motions, and the parallax.
		}
		\label{fig:vet}
	\end{center}
\end{figure*}

@@

The mosaic in Figure~\ref{fig:cqvs} shows phased light curves for the
\ngoods\ CQVs.  The objects are sorted first in order of the number of
TESS 120-second cadence sectors in which they clearly demonstrated the
CQV phenomenon, and secondarily by descending brightness.  The top
five objects by this metric are TIC~300651846 ($T$$=$13.5, 12
sectors); TIC~402980664 ($T$$=$11.1, 7 sectors); TIC~89463560
($T$$=$13.5, 5 sectors); TIC~363963079 ($T$$=$12.9, 5 sectors); and
TIC~294328887 ($T$$=$14.2, 4 sectors).  The brightest five CQVs span
9.3$<$$T$$<$11.1; the faintest five span 14.5$<$$T$$<$15.0.  The
fastest five have periods spanning 3.6\,hr$<$$P$$<$6.2\,hr, and the
slowest five span 27\,hr$<$$P$$<$38\,hr.  Regarding likely astrometric
or visual binaries, \ngoodhighruwe/\ngoods\ high-quality CQVs have a
Gaia DR2 RUWE$>$2; \nmaybehighruwe/\nmaybes\ of the ambiguous sample
share this characteristic.\footnote{For an explanation of the renormalized unit weight error (RUWE), see the GAIA DPAC technical note \url{http://www.rssd.esa.int/doc_fetch.php? id=3757412}.  Elevated RUWEs imply elevated noise relative to a single-source astrometric model. This can be caused by marginally resolved point sources, photometric variability, or astrometric motion.}

@@

@@

\subsection{Binaries, massive CQVs, and a new class of false positive}

The high-RUWE objects in Figure~\ref{fig:catalogscatter} tend to be
among the more massive half of the CQV sample.  This suggests that either
that the RUWE indicator is not sensitive to binarity among lower mass
stars, or that these ``high-mass CQVs'' may be unresolved blends in
which the lower mass companion is produces the CQV signal.

@@

The two most massive CQVs in our sample present a relevant case study.
To date, the only stars reported to show the CQV phenomenon are M
dwarfs, with typical stellar masses $\lesssim$0.3\,$M_\odot$
\citep{2017AJ....153..152S,2022AJ....163..144G}.  TIC~405754448 and
TIC~405910546 however appear to have masses of
$\approx$0.82\,$M_\odot$ and $\approx$0.60\,$M_\odot$ respectively.
The next-highest masses in our sample are $\approx$0.40\,$M_\odot$.
The masses for the former two objects are consistent with their
locations in the color--absolute magnitude diagram, and with their
membership in LCC.  However the light curve morphology of
TIC~405910546 is unique among the CQV sample, and suggestive of an
eclipsing binary; it should be studied further to clarify its
classification.  For TIC~405754448, we expect that this source {\it
is} an unresolved binary, because RUWE$_{\rm DR2}$=4.7, and because
the raw light curves in Sectors 11, 37, and 38 show an
additional $\approx$6.5\,day, $\approx$0.3\% amplitude sinusoidal
signal suggestive of an unresolved photometric companion.  If
TIC~405754448 is a K+M binary, then the flux ratio between the primary
and secondary would be expected to be $\approx$10:1.  It would be
challenging for the M dwarf to produce the observed 12.9\,hr signal
with such a flux deficit, but not impossible.  While both of
these objects suggest that the CQV phenomenon may extend up in mass to
pre-main-sequence K dwarfs, more data are needed to fully substantiate
this claim.

@@

An independent set of CQVs exhibits small dips roughly half a cycle
after large dips; it is tempting to label the small dips secondary
eclipses.  Such sources include TIC~402980664 (Sector 25; relative to
the sharpest, deepest minimum), TIC~89463560 (Sector 36),
TIC~224283342 (Sector 2), and TIC~442571495 (Sector 12).

It is challenging to interpret the significance of such apparent
regularities;  the objects exhibit such a wide range of variability
that for seemingly any ``regularity'' one might posit, it is easy to
come up with counter-example objects which do not follow the trend.
In other words, these half-cycle CQVs could simply be a product of the
human tendency to pattern match; alternatively, they might yield some
important physical significance.

@@

\begin{enumerate}[leftmargin=*]
	%
    \item CQVs are quasiperiodic.  The mean periods remain fixed over
      the $>$1{,}000-cycle baseline of available observations; but the
      light curve shapes always evolve (Figure~\ref{fig:evoln}).
  %
    \item The same CQV can show multiple periods simultaneously.  LP
      12-502, for instance, showed dips with four distinct periods
      within $\pm 0.3\%$ of its fundamental period, sometimes
      simultaneously, and each lasting for up to 50 cycles
      (Figure~\ref{fig:lpriver0}).
  %
    \item CQVs evolve over timescales that are both secular
      ($>$100\,cycles) and impulsive ($<$1 cycle).  Dip growth seems
      to happen over durations of at least ten cycles, and slow dip
      decay can also occur.  ``State-switches'' correspond to dips to
      collapsing instantaneously, and are almost always linked with
      observed optical flares.  This suggests that the occulting
      material is sensitive to sudden changes in the magnetic field.
  %
    \item The rate of dip evolution can be used to place a
      model-dependent constraint on how much material is either being
      accreted or ejected during the state changes
      (Section~\ref{subsec:massflux}).  Order of magnitude estimates
      require at least an asteroid belt's worth of dust
      ($10^{-4}$\,$M_\oplus$) over $10^8$ years, or $\approx$$10^{-2}$\,$M_\oplus$
      if the occulting material is gas.
  %
    \item The CQV phenomenon persists for $\gtrsim$150\,Myr, based on
      the existence of multiple CQVs in AB Dor, the Pleiades, and
      Psc-Eri (Section~\ref{subsec:starprops}).  It may even extend to
      200\,Myr, based on the one CQV we found in the Carina Near
      moving group (TIC~294328887; $\approx$200\,Myr).  However the
      lack of detected CQVs in the Hyades and Praesepe suggests that
      the lifetime of the phenomenon is limited to the first few
      hundred million years.
	%
    \item The duty cycle for CQVs seems to be $\approx$$75$\%, based
      on the fraction of bonafide CQVs that either turned on or turned
      off during TESS re-observations, two years after the initial
      observation.
  %
    \item While the majority of detected CQVs have masses of
      0.1-0.5\,$M_\odot$, the upper mass limit for the CQV phenomenon
      might extend into the K dwarf regime ($<$$0.8$\,$M_\odot$).
      TIC~405754448 and TIC~405910546, with masses of $\approx$0.8 and
      $\approx$0.6\,$M_\odot$, are the two massive outliers.
      However both have peculiarities (Section~\ref{subsec:massive}), and 
      we encourage additional
      scrutiny of these objects in future work.
  %
    \item Surprising analogs to the CQVs exist in magnetic B stars,
      but only those with multipolar field topologies.  Since most
      magnetic B stars have dipolar magnetic fields, this suggests
      that the CQV dips and warps are similarly being sculpted by the
      stellar magnetic fields, and that the magnetic fields themselves
      are probably also multipolar.
  %
    \item The closest CQVs to the Sun are at distances of 15-20\,pc,
      and the brightest have $V$$\approx$12 ($J$$\approx$7.5).  We
      have found most of them in this work, since our CQV sample was
      $\gtrsim$80\% complete within 30\,pc.  The lack of CQVs in the
      volume-complete $<$15\,pc sample of 0.1-0.3\,$M_\odot$ stars
      analyzed by \citet{2021AJ....161...63W} is consistent with this
      estimate.  Expanding our analysis of the TESS data to the full
      frame images would yield a truly volume-limited selection
      function, and would expand the CQV census by about a factor of
      two within 50\,pc, and by a factor of ten within 100\,pc.
\end{enumerate}

>>>>

\begin{figure*}[!tp]
	\begin{center}
		\centering
		\includegraphics[width=0.96\textwidth]{f2.pdf}
		\vspace{-0.3cm}
		\caption{
			{\bf Validation plots used to  label CQVs}.  The complete figure
			set, with one image per sector for each of the \nallcands\
      objects in Table~\ref{tab:thetable}
			is available online {\bf For internal collaboration review:
				\url{https://www.dropbox.com/sh/khtwk5a6z0zgrif/AABRG21Ire7VS4BUnaD1hOR8a?dl=0}}.
			Panels are as follows.
			{\it a)}: Phase-folded light curve; gray points are raw 2-minute
			data and black points are binned to 200 points per cycle.
			{\it b)}: Phase-dispersion minimization (PDM) periodogram.
			Dotted lines show up to the 10$^{\rm th}$ harmonic and
			subharmonic.
			{\it c)}: DSS finder chart, with 1- and 2-TESS pixel radius
			circles displayed for scale.
			{\it d)}: Cleaned light curve, binned to 20-minute cadence, in
			Barycentric TESS Julian Date (BTJD).
			{\it e)}: Phase-folded light curve, binned to 100 points per
			cycle.  The gray line denotes the automated spline-fit to the
			wrapped phase-folded light curve, and small gray triangles
			denote automatically identified local minima.
			{\it f)}: Phase-folded light curve at twice the peak period.
			{\it g)}: Phase-folded light curve at half the peak period.
			{\it h)}: Phase-folded time-series within the ``background''
			aperture defined in the SPOC light curves.
			{\it i)}: Phase-folded flux-weighted centroid in the column
			direction.
			{\it j)}: Phase-folded flux-weighted centroid in the row
			direction.
			{\it k)}: Gaia DR2 color--absolute magnitude diagram.     
			The gray background denotes stars within 100\,pc.
			{\it l)}: Information from Gaia DR2, TIC8, and the automated
			dip-counting search pipeline.  ``Neighbors'', abbreviated
			``nbhr'', are listed within apparent distances of 2 TESS pixels
			if $\Delta T$$<$2.5.
			{\it m)}: BANYAN-$\Sigma$ v1.2 association probabilities, calculated
			using positions, proper motions, and the parallax.
		}
		\label{fig:vet}
	\end{center}
\end{figure*}

@@

The mosaic in Figure~\ref{fig:cqvs} shows phased light curves for the
\ngoods\ CQVs.  The objects are sorted first in order of the number of
TESS 120-second cadence sectors in which they clearly demonstrated the
CQV phenomenon, and secondarily by descending brightness.  The top
five objects by this metric are TIC~300651846 ($T$$=$13.5, 12
sectors); TIC~402980664 ($T$$=$11.1, 7 sectors); TIC~89463560
($T$$=$13.5, 5 sectors); TIC~363963079 ($T$$=$12.9, 5 sectors); and
TIC~294328887 ($T$$=$14.2, 4 sectors).  The brightest five CQVs span
9.3$<$$T$$<$11.1; the faintest five span 14.5$<$$T$$<$15.0.  The
fastest five have periods spanning 3.6\,hr$<$$P$$<$6.2\,hr, and the
slowest five span 27\,hr$<$$P$$<$38\,hr.

@@

\subsection{Binarity}
\label{subsec:binarity}

The main types of binarity of interest in this work were those that
could confuse our understanding of CQVs.  For instance, within the
21$''$ TESS pixels, binary companions can produce multiple photometric
signals and hinder our ability to correctly identify the star hosting
the CQV signal.  Binary companions can also bias photometric magnitude
and color measurements, thereby biasing our stellar parameter
estimates.  To attempt to identify binaries, we considered the
following lines of information.
%
\paragraph{Radial velocity scatter}---We examined diagrams of the Gaia
DR3 ``radial velocity error'' as a function of stellar color for all
\ncqvsnodebunked\ CQVs and candidate CQVs.  Since this quantity
represents the standard deviation of the non-published Gaia RV time
series, outliers can suggest single-lined spectroscopic binarity.
These plots showed two clusters of stars, at $\lesssim$10\,\kms and
20-25\,\kms.  We there adopted a threshold of $20$\,km\,s$^{-1}$
to flag possible single-lined spectroscopic binaries, which selected
\nrvscatter\ stars: TIC~405910546, TIC~224283342, and TIC~280945693.
%
\paragraph{RUWE}---We examined plots of Gaia DR3 renormalized unit
weight error (RUWE) as a function of brightness as color.\footnote{For
an explanation of the renormalized unit weight error (RUWE), see the
GAIA DPAC technical note \url{http://www.rssd.esa.int/doc_fetch.php?
id=3757412}.}  Elevated RUWEs imply elevated noise relative to a
single-source astrometric model.  This can be caused by marginally
resolved point sources skewing the single-transit centroid
measurements,  intrinsic photometric variability, or intrinsic
astrometric motion.  Based on this exercise, we adopted a threshold of
RUWE$_{\rm DR3}$ $>$ 2 to flag sources with excess astrometric noise.
This threshold yielded \ngoodhighruwe/\ngoods\ high-quality CQVs with
a Gaia DR2 RUWE$>$2; \nmaybehighruwe/\nmaybes\ of the ambiguous sample
share this characteristic.
%
\paragraph{Gaia DR3 non-single stars}---Gaia DR3 included a
``non\_single\_star'' column that flagged known or suspected
eclipsing, astrometric, and spectroscopic binaries.  None of the stars
in our CQV sample were identified as possible binaries in this column.
%
\paragraph{Multiple periodic TESS signals}---During our visual
analysis of the TESS light curves and PDM periodograms, we flagged
sources with beating light curves, and with PDM periodograms that
showed multiple periods.  We also attempted to disentangle the two
signals by subtracting the mean CQV signal over each sector, and then
repeating the phase-dispersion minization analysis.  The results of
this effort are summarized in the secondary periods, $P_{\rm sec}$,
listed in Table~\ref{tab:thetable}.  {\bf We found additional periods
in XX of the CQVs, of which YY also had elevated RUWEs}.

%FIXME FIXME FIXME:
% Need to actually incorporate into the table!

We incorporated each of these sources of information into a column in
Table~\ref{tab:thetable} that expresses whether any given source
passed each of the binarity indicators.
We discuss some important--and complicated--aspects of the systems
that seem to be binary in Section~\ref{subsec:discbinary}.

@@

\subsection{Binaries, massive CQVs, and a new class of false positive}
\label{subsec:discbinary}

%FIXME FIXME FIXME
%TODO: update to give some flavor for how many sources show second periods!

The high-RUWE objects in Figure~\ref{fig:catalogscatter} tend to be
among the more massive half of the CQV sample.  This suggests that
some of the ``high-mass CQVs'' may be unresolved blends in which the
lower mass companion produces the CQV signal.

@@

The two most massive CQVs in our sample, nominally in the K dwarf mass
regime, present a relevant case study.  To date, the only stars
reported to show the CQV phenomenon are M dwarfs, with typical stellar
masses $\lesssim$0.3\,$M_\odot$
\citep{2017AJ....153..152S,2022AJ....163..144G}.  TIC~405754448 and
TIC~405910546 appear to have masses of $\approx$0.82\,$M_\odot$ and
$\approx$0.60\,$M_\odot$ respectively.  The next-highest masses in our
sample are $\approx$0.40\,$M_\odot$.

Based on their locations locations in the color--absolute magnitude
diagram and membership in LCC, both of the K dwarfs do indeed appear
to be high mass.  However in detail, both are subject to ambiguities
in interpretation.  The light curve morphology of TIC~405910546 is
unique in the CQV sample, and suggestive of an eclipsing binary.
Independently, TIC~405910546 was one of the \nrvscatter\ CQVs flagged
with a Gaia DR3 radial velocity scatter exceeding 20\,\kms.  Combined,
these factors suggest that TIC~405910546 might be a pre-main-sequence
eclipsing binary; it should be studied further to clarify this
classification.  For TIC~405754448, we believe that this source {\it
is} an unresolved binary, because RUWE$_{\rm DR2}$=4.7, and because
the raw light curves in Sectors 11, 37, and 38 show an additional
$\approx$6.5\,day, $\approx$0.3\% amplitude sinusoidal signal
suggestive of an unresolved photometric companion.  If TIC~405754448
is a K+M binary, then the flux ratio between the primary and secondary
would be expected to be $\approx$10:1.  It would be challenging for
the M dwarf to produce the observed 12.9\,hr signal with such a flux
deficit, but not impossible.  While both of these objects suggest that
the CQV phenomenon may extend up in mass to pre-main-sequence K
dwarfs, more data are needed to fully substantiate this claim.

@@

An independent set of CQVs exhibits small dips roughly half a cycle
after large dips; it is tempting to label the small dips secondary
eclipses.  Such sources include TIC~402980664 (Sector 25; relative to
the sharpest, deepest minimum), TIC~89463560 (Sector 36),
TIC~224283342 (Sector 2), and TIC~442571495 (Sector 12).
For the particular case of TIC~224283342, this observation, coupled
with Gaia DR3 radial velocities suggesting a radial
velocity scatter exceeding 20\,\kms, suggests that the system could be
a low-mass eclipsing binary transiting a giant polar spot, analogous to
TOI-3884 \citep{2022A&A...667L..11A}.

In general however, it is challenging to interpret the significance of
the potential regularities listed above;  the CQVs exhibit such a wide
range of variability that for seemingly any ``regularity'' one might
posit, it is easy to come up with counter-example objects which do not
follow the trend.  In other words, these half-cycle CQVs could simply
be a product of the human tendency to pattern match.  The alternative
is that some subset of these pattern could  have real physical
significance.

@@

\begin{enumerate}[leftmargin=*]
	%
    \item CQVs are quasiperiodic.  The mean periods remain fixed over
      the $>$1{,}000-cycle baseline of available observations; but the
      light curve shapes always evolve (Figure~\ref{fig:evoln}).
  %
    \item The same CQV can show multiple periods simultaneously.  LP
      12-502, for instance, showed dips with four distinct periods
      within $\pm 0.3\%$ of its fundamental period, sometimes
      simultaneously, and each lasting for up to 50 cycles
      (Figure~\ref{fig:lpriver0}).
  %
    \item CQVs evolve over timescales that are both secular
      ($>$100\,cycles) and impulsive ($<$1 cycle).  Dip growth seems
      to happen over durations of at least ten cycles, and slow dip
      decay can also occur.  ``State-switches'' correspond to dips to
      collapsing instantaneously, and are almost always linked with
      observed optical flares.  This suggests that the occulting
      material is sensitive to sudden changes in the magnetic field.
  %
    \item The rate of dip evolution can be used to place a
      model-dependent constraint on how much material is either being
      accreted or ejected during the state changes
      (Section~\ref{subsec:massflux}).  Order of magnitude estimates
      require at least an asteroid belt's worth of dust
      ($10^{-4}$\,$M_\oplus$) over $10^8$ years, or $\approx$$10^{-2}$\,$M_\oplus$
      if the occulting material is gas.
  %
    \item The CQV phenomenon persists for $\gtrsim$150\,Myr, based on
      the existence of multiple CQVs in AB Dor, the Pleiades, and
      Psc-Eri (Section~\ref{subsec:starprops}).  It may even extend to
      200\,Myr, based on the one CQV we found in the Carina Near
      moving group (TIC~294328887; $\approx$200\,Myr).  However the
      lack of detected CQVs in the Hyades and Praesepe suggests that
      the lifetime of the phenomenon is limited to the first few
      hundred million years.
	%
    \item The duty cycle for CQVs seems to be $\approx$$75$\%, based
      on the fraction of bonafide CQVs that either turned on or turned
      off during TESS re-observations, two years after the initial
      observation.
  %
    \item The majority of detected CQVs have masses of
      0.1-0.5\,$M_\odot$.  Two sources, TIC~405754448 and
      TIC~405910546, have masses that appear to exceed the M dwarf
      limit.  However both are potentially binaries, and this may
      confuse our ability to accurate identify the source of the CQV
      signal (Section~\ref{subsec:massive}).  We encourage additional
      scrutiny of these objects in future work.
  %
    \item Surprising analogs to the CQVs exist in magnetic B stars,
      but only those with multipolar field topologies.  Since most
      magnetic B stars have dipolar magnetic fields, this suggests
      that the CQV dips and warps are similarly being sculpted by the
      stellar magnetic fields, and that the magnetic fields themselves
      are probably also multipolar.
  %
    \item The closest CQVs to the Sun are at distances of 15-20\,pc,
      and the brightest have $V$$\approx$12 ($J$$\approx$7.5).  We
      have found most of them in this work, since our CQV sample was
      $\gtrsim$80\% complete within 30\,pc.  The lack of CQVs in the
      volume-complete $<$15\,pc sample of 0.1-0.3\,$M_\odot$ stars
      analyzed by \citet{2021AJ....161...63W} is consistent with this
      estimate.  Expanding our analysis of the TESS data to the full
      frame images would yield a truly volume-limited selection
      function, and would expand the CQV census by about a factor of
      two within 50\,pc, and by a factor of ten within 100\,pc.
\end{enumerate}
